---
title: "Stat 433 Week 3"
author: "Alex Hertadi"
date: "9/27/2021"
output: html_document
---

```{r}
library(dplyr)

# install.packages("nycflights13")
library(nycflights13)
library(ggplot2)
library(lubridate)
```

1.
```{r}
flights %>% 
  count(is.na(dep_time))
```
8255 flights have a missing dep_time
```{r}
mis = flights %>% filter(is.na(dep_time))
mis
notail = flights %>% filter(is.na(tailnum))
notail
```
The entries with a missing dep_time are also missing dep_delay, arr_time, arr_delay, and air_time. These entries might represent flights that got cancelled.
2512 of the entries also do not have tailnumbers, indicating that there might've not been an aircraft assigned to these schedules

2.
```{r}
#isolated table without NA entries
noCancel = flights %>% filter(!is.na(dep_time), !is.na(sched_dep_time))
noCancel %>% select(dep_time,sched_dep_time,dep_delay)
noCancelTransform = noCancel %>% 
        mutate(sched_dep_time = 60*hour + minute) %>% 
        mutate(dep_time = sched_dep_time + dep_delay)
noCancelTransform

```
3.
```{r}
flights %>% 
  group_by(day = yday(time_hour)) %>% 
  summarise(
    delay = mean(arr_delay, na.rm = TRUE),
    canceled = sum(is.na(dep_time))
  ) %>%
  filter(canceled<300) %>% 
  ggplot(mapping = aes(y = canceled, x = delay)) +
  geom_point()
  
```

It seems that there exists a relationship between the average delay and number of flight cancellations especially at a certain threshold. Outliers with cancellation numbers above 300 are removed for a clearer view of the graph.  
